---
title: "India Human Development Survey (IHDS)"
output:
  rmdformats::html_clean:
    code_folding: hide
    fig_width: 8
    fig_height: 6
    highlight: kate
    thumbnails: false
    lightbox: true
    gallery: false
    self_contained: false
---

```{r knitr_init, echo=FALSE, warning=FALSE, message=FALSE, results="asis"}
#devtools::install_github("hathawayj/tidyICPSR")
#install.packages("pacman")
#devtools::install_github("hafen/geofacet", ref = "feature-hexmapr")
library(knitr)
## Global options
options(max.print = "75")
opts_chunk$set(
  echo = TRUE,
  # cache = TRUE,
  prompt = FALSE,
  tidy = FALSE,
  comment = NA,
  message = FALSE,
  warning = FALSE)
opts_knit$set(width = 75)
```

# Introduction

## Visualization

The visualizations below provide an example of using [R](https://www.r-project.org/){target="blank"} to summarize and visualize anthropometric data and other household data that was collected in India during the [IHDS study](https://www.ihds.umd.edu/){target="blank"}.  The scripts used to get and create the data for the visualizaings below can be found [here](https://github.com/hathawayj/ihds/tree/master/scripts){target="blank"}. The IHDS can be downloaded from [the ICPSR](https://www.icpsr.umich.edu/icpsrweb/DSDR/series/507){target="blank"}.

- **[data download and format scripts](https://github.com/hathawayj/ihds/tree/master/scripts){target="blank"}**

## tidyICPSR

We have developed an R package called [tidyICPSR](https://github.com/hathawayj/tidyICPSR){target="blank"} that provides functionality to get the two studies downloaded and formatted into a [tidy format](http://garrettgman.github.io/tidying/){target="blank"} where all column values are transformed into correct variable values.  The full column names can be used as well. The package is still in early development and is currently meant more for a proof of concept and could be helpful for those that would like to use R with the data from ICPSR.

- **[Rmd file that created this page](https://hathawayj.github.io/ihds/index.Rmd)**
- **Use the' 'Code' buttons to see the code to generate the graphic**

# 2011 IHDS Exploration (Individual)

```{r loaddata}
#devtools::install_github("hathawayj/tidyICPSR")
#install.packages("pacman")
pacman::p_load(knitr, rmdformats, tidyverse, ggthemes, forcats, geofacet, tidyICPSR, growthstandards)
ind11 <- read_rds(path = "../data/artifacts/2011_36151_India.Rds")


### Build data items for plots
### 

dat <- ind11 %>%
  mutate(stateid = str_remove_all(stateid, " [:digit:]+") %>% 
           str_replace_all("&", "and") %>% str_replace_all("Orissa", "Odisha"),
         ro3 = str_remove_all(ro3, " [:digit:]+"), ro3 = str_remove_all(ro3, ";")) %>%
  select(idperson, stateid, hhid, distid, psuid, hhsplitid, idhh, 
         anthro_id = ap3, age = ro5, height1 = ap5, height2 = ap6, 
         anthro_position = ap7, weight1 = ap8, weight2 = ap9, 
         sex = ro3, relation_head = ro4, diarrhea = sm7, 
         hours_work_year = wkhours, annual_wages = wsearn)

household <- dat %>%
  group_by(hhid, stateid, distid, psuid, hhsplitid, idhh) %>%
  summarise(n = n(), children = sum(age <= 18, na.rm = TRUE), 
            household_wages = sum(annual_wages, na.rm = TRUE),
            household_hours = sum(hours_work_year, na.rm = TRUE)) %>%
  ungroup()


kids <- dat %>%
  filter(age <= 5) %>% 
  left_join(household)



```


## Household Income

The household income numbers are close to the other reports we found. The median household income from this study data is `r as.character(round(median(household$household_wages),0))` NRP.

 - https://cmie.com/kommon/bin/sr.php?kall=warticle&dt=2016-07-04%2013:45:29&msec=170
 - https://www.quora.com/What-is-the-median-income-of-households-in-different-states-of-India

The graphic below shows the median (dot) and the 25th and 75th percentiles (linerange) of household income for each region of India.

```{r income_viz_box, fig.height=5}


household %>%
  mutate(stateid = fct_reorder(stateid, household_wages, median)) %>% 
  group_by(stateid) %>%
  summarise(p25 = quantile(household_wages, .25), median = median(household_wages), p75 = quantile(household_wages, .75)) %>%
  ungroup() %>%
  ggplot(aes(y = median/10000, x = stateid)) +
  geom_pointrange(aes(ymin = p25/10000, ymax = p75/10000)) +
# geom_vline(aes(xintercept = median), data = state_labels) +
  theme_bw() +
  labs(y = "Household income (10,000 INR)", x = "Region", title = "Distribution of Household Income") +
  theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))  

```

Below we have displayed the household income (x-axis), and the line shows the proportion of households from the study at a given income. As of this work [100,000 INR converts to ~$1500](https://www.google.com/search?q=100000+INR+to+dollars&oq=100000+INR+to+dollars&aqs=chrome..69i57.6348j1j9&sourceid=chrome&ie=UTF-8){target="blank"}. 

```{r income_viz, fig.height=8}

f90 <- function(x){ quantile(x, .9, na.rm = TRUE)}

state_labels <- household %>% 
  mutate(stateid = fct_reorder(stateid, household_wages, f90)) %>%
  group_by(stateid) %>%
  summarise(sum_n = sum(n), mean = mean(household_wages, na.rm = TRUE), 
            median = median(household_wages, na.rm = TRUE), 
            p90 = quantile(household_wages, .9, na.rm = TRUE),
            p10 = quantile(household_wages, .1,na.rm = TRUE)) %>%
  ungroup()


household %>%
  mutate(stateid = fct_reorder(stateid, household_wages, f90)) %>%
  ggplot(aes(x = household_wages/100000)) +
  # geom_rect(data = state_labels, aes(fill = p90),
  #   xmin = -Inf, ymin = -Inf, xmax = Inf, ymax = Inf, alpha = 0.5,
  #   inherit.aes = FALSE) +
  stat_ecdf(geom = "step", pad = TRUE) +
  theme_bw() +
  coord_cartesian(xlim = c(0, 10)) +
  scale_x_continuous(breaks = seq(0, 10, by = 2)) +
  scale_y_continuous(breaks = seq(0, 1, by = .2)) +
  geom_hline(color = "darkgrey", yintercept = c(.9)) +
  geom_vline(data = state_labels, aes(xintercept = p90/100000), color = "darkgrey") +
  labs(x = "Household income (100,000 INR)", y = "Proportion", title = "Distribution of Household Income") +
  facet_wrap(~stateid, labeller = labeller(stateid = label_wrap_gen(10)), nrow = 6) 

#  facet_geo(~stateid, grid = india_grid2, labeller = labeller(stateid = label_wrap_gen(10))) 

```


## Household Size

The household count numbers seem to make sense based on numbers reported at the following links. The median household size from this study data is `r round(median(household$n),2)`. There is not much exciting in the below graphic. 

- https://en.wikipedia.org/wiki/Indian_states_ranking_by_household_size
- http://www.arcgis.com/home/item.html?id=6cf22970ea8c4b338a196879397a76e4



```{r size_viz_box, fig.height=5}
household %>%
  mutate(stateid = fct_reorder(stateid, n, median)) %>% 
  group_by(stateid) %>%
  summarise(p25 = quantile(n, .25), median = median(n), p75 = quantile(n, .75)) %>%
  ungroup() %>%
  ggplot(aes(y = median, x = stateid)) +
  geom_pointrange(aes(ymin = p25, ymax = p75)) +
# geom_vline(aes(xintercept = median), data = state_labels) +
  theme_bw() +
  labs(y = "Number in household", x = "Region", title = "Distribution of Household Size") +
  scale_y_continuous(breaks = seq(0,30, by = 5), limits = c(0, 15)) +
  theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))  

```

## One to four years old height distributions

Some data quality issues will require further investigation for the heights. We have converted the heights into Z-scores (HAZ) using `growthstandards::who_htmc2zscore()`. The ages are grouped by year in the data so the HAZ values will be much more variable than usual as some of the values could be off by as much as 364 days. 

```{r, fig.height=8}
# Household heights by age
kids %>%
  filter(age < 5, age > 0) %>%
  mutate(age = ifelse(age == 0, .5, age),
    zscore = who_htcm2zscore(agedays = .$age*365.25 , htcm = .$height2, sex = .$sex ), 
    stateid = fct_reorder(stateid, zscore, median, na.rm = T)) %>%
  ggplot(aes(y = zscore, x = factor(age)), fill = "darkgrey") +
  geom_jitter(height = 0, width = .25, alpha = .2) +
  geom_boxplot(fill = NA, color = "limegreen", outlier.colour = NA) +
  theme_bw() +
  scale_color_brewer(type = "qual") +
  coord_cartesian(ylim = c(-7, 7)) +
  labs(y = "Height (ft)", x = "Age", title = "Distribution of height by age group") +
  facet_wrap(~stateid)

```


## 0-1-year-old heights by income category

Since the HAZ values are very sensitive to incorrect age day values, we show the heights of those labeled as zero years old. Remember that this figure shows all newborns from 0-365 days.

```{r, fig.width=6}
# Household heights by age
kids %>%
  filter(age == 0) %>%
  mutate(stateid = fct_reorder(stateid, height2, median, na.rm = T)) %>%
  ggplot(aes(y = height2, x = stateid), fill = "darkgrey") +
  geom_jitter(height = 0, width = .25, alpha = .2) +
  geom_boxplot(fill = NA, color = "limegreen", outlier.colour = NA) +
  theme_bw() +
  scale_color_brewer(type = "qual") +
  labs(y = "Height (cm)", x = "Region", title = "< 1 year heights") +
  coord_flip()
```


```{r, fig.height=8, fig.width = 6, include=FALSE}
# Household heights by age
kids %>%
  filter(age < 5, age > 0) %>%
  mutate(zscore = who_htcm2zscore(agedays = .$age*365.25 , htcm = .$height2, sex = .$sex )) %>%
  group_by(stateid) %>%
  summarize(median_haz = median(zscore, na.rm = TRUE),
            p90_wages = quantile(household_wages, .9,  na.rm = TRUE),
            n = sum(n)) %>%
  ungroup() %>%
  filter(n > 500) %>%
  ggplot(aes(x = p90_wages/100000, y = median_haz)) +
  geom_point() +
  theme_bw() +
  labs(y = "Medain HAZ", x = "90th percentile household wage", title = "") 

```



<!-- ## Under 5 years old weight distributions -->

<!-- Some data quality issues will require further investigation. -->

<!-- ```{r, fig.height=16, fig.width = 12} -->
<!-- kids %>% -->
<!--   filter(age <= 5) %>% -->
<!--   ggplot(aes(y = weight2 * 2.20462, x = factor(age)), fill = "darkgrey") + -->
<!--   geom_boxplot() + -->
<!--   #  geom_jitter(aes(color = sex), alpha = .5, width = .25) + -->
<!--   facet_geo(~stateid, grid = india_grid2) + -->
<!--   theme_bw() + -->
<!--   scale_color_brewer(type = "qual") + -->
<!--   coord_cartesian(ylim = c(0, 45)) + -->
<!--   labs(y = "weight (lbs)", x = "gender", color = "gender", title = "Distribution of height by age group") -->


<!-- ``` -->

<!-- ## 0-1-year-old weights by income category -->

<!-- Remember that this figure shows all newborns from 0-12 months. -->

<!-- ```{r, fig.height=16, fig.width = 12} -->
<!-- kids %>% -->
<!--   filter(age <= 1) %>% -->
<!--   mutate(income_groups =  -->
<!--            case_when( household_wages/ 10000 <= 2.5 ~ "less than 25k", -->
<!--                       household_wages/ 10000 <= 5 ~ "between 25k & 50k", -->
<!--                       household_wages/ 10000 <= 10 ~ "between 50k & 100k", -->
<!--                       household_wages/ 10000 <= 25 ~ "between 100k & 250k", -->
<!--                       household_wages/ 10000 > 25 ~ "greater than 250k")) %>% -->
<!--   ggplot(aes(y = weight2 * 2.200462, x = income_groups)) + -->
<!--   scale_y_continuous(limits = c(0, 25)) +  -->
<!--   geom_boxplot() + -->
<!--   labs(x = "Household wages (INR)", y = "weight (lbs)") + -->
<!--   coord_flip() + -->
<!--   theme_bw() -->
<!-- ``` -->



<!-- ```{r, fig.height=16, fig.width = 12} -->
<!-- kids %>% -->
<!--   filter(age <= 1) %>% -->
<!--   mutate(income_groups =  -->
<!--            case_when( household_wages/ 10000 <= 2.5 ~ "less than 25k", -->
<!--                       household_wages/ 10000 <= 5 ~ "between 25k & 50k", -->
<!--                       household_wages/ 10000 <= 10 ~ "between 50k & 100k", -->
<!--                       household_wages/ 10000 <= 25 ~ "between 100k & 250k", -->
<!--                       household_wages/ 10000 > 25 ~ "greater than 250k")) %>% -->
<!--   ggplot(aes(y = weight2 * 2.200462, x = income_groups)) + -->
<!--   scale_y_continuous(limits = c(0, 25)) +  -->
<!--   geom_boxplot() + -->
<!--   facet_geo(~stateid, grid = india_grid2) + -->
<!--   labs(x = "Household wages (INR)", y = "weight (lbs)") + -->
<!--   coord_flip() + -->
<!--   theme_bw() -->
<!-- ``` -->


# 2011 IHDS Exploration (Births)

## Distribution of number of births per household

```{r, fig.width = 6}
dat <- read_rds(path = "../data/artifacts/2011_36151_India_births.Rds")


dat %>%
  group_by(idhh, stateid, mother_id) %>%
  summarise(births = max(nbirths)) %>%
  ungroup() %>%
  mutate(stateid = fct_reorder(stateid, births, quantile, probs = .75)) %>%
  ggplot(aes(x = stateid, y = births)) +
  geom_boxplot() + 
  coord_flip() +
  theme_bw() +
  labs(y = "Births per Mother", x = "Region")
```



# IHDS Data Prep

The development R package [tidyICPSR]() has the functions used to download and prep the data.

# Code

All of the R code to download, process, and replicate this analysis (including the code that generates this document) is available [here](https://github.com/hathawayj/ihds).




