---
title: "India Human Development Survey (IHDS)"
output:
  rmdformats::html_clean:
    fig_width: 8
    fig_height: 6
    highlight: kate
    thumbnails: false
    lightbox: true
    gallery: false
    self_contained: false
---

```{r knitr_init, echo=FALSE, warning=FALSE, message=FALSE, results="asis"}
#devtools::install_github("hathawayj/tidyICPSR")
#install.packages("pacman")
pacman::p_load(knitr, rmdformats, tidyverse, ggthemes, forcats, geofacet, tidyICPSR)

## Global options
options(max.print = "75")
opts_chunk$set(
  echo = FALSE,
  # cache = TRUE,
  prompt = FALSE,
  tidy = FALSE,
  comment = NA,
  message = FALSE,
  warning = FALSE)
opts_knit$set(width = 75)
```

# Introduction

The IHDS is summarized [here](https://www.ihds.umd.edu/){target="blank"} and can be downloaded from [the ICPSR](https://www.icpsr.umich.edu/icpsrweb/DSDR/series/507){target="blank"}.  We have developed an R package called [tidyICPSR](https://github.com/hathawayj/tidyICPSR){target="blank"} that provides functionality to get the two studies downloaded and formatted into a [tidy format](http://garrettgman.github.io/tidying/){target="blank"} with all the column names and column values correct. The package is still in early development and is currently meant more for a proof of concept and could be helpfull for those that would like to use R with the data from ICPSR.

Note that the IHDS study collected data on subjects from the Union territories of India.  They are not shown in the spatial graphics below. 

# 2011 IHDS Exploration (Individual)

```{r loaddata}

ind11 <- read_rds(path = "../data/artifacts/2011_36151_India.Rds")


### Build data items for plots
### 

dat <- ind11 %>%
  mutate(stateid = str_remove_all(stateid, " [:digit:]+") %>% 
           str_replace_all("&", "and") %>% str_replace_all("Orissa", "Odisha"),
         ro3 = str_remove_all(ro3, " [:digit:]+")) %>%
  select(idperson, stateid, hhid, distid, psuid, hhsplitid, idhh, 
         anthro_id = ap3, age = ro5, height1 = ap5, height2 = ap6, 
         anthro_position = ap7, weight1 = ap8, weight2 = ap9, 
         sex = ro3, relation_head = ro4, diarrhea = sm7, 
         hours_work_year = wkhours, annual_wages = wsearn)

household <- dat %>%
  group_by(hhid, stateid, distid, psuid, hhsplitid, idhh) %>%
  summarise(n = n(), children = sum(age <= 18, na.rm = TRUE), 
            household_wages = sum(annual_wages, na.rm = TRUE),
            household_hours = sum(hours_work_year, na.rm = TRUE)) %>%
  ungroup()


kids <- dat %>%
  filter(age <= 5) %>% 
  left_join(household)

```


## Household Income

The household income numbers are close to the reports I could find. The median household income from this study data is `r as.character(round(median(household$household_wages),0))` NRP.

 - https://cmie.com/kommon/bin/sr.php?kall=warticle&dt=2016-07-04%2013:45:29&msec=170
 - https://www.quora.com/What-is-the-median-income-of-households-in-different-states-of-India

```{r income_values}
state_labels <- household %>% 
  group_by(stateid) %>%
  summarise(sum = sum(n), mean = mean(household_wages, na.rm = TRUE), median = median(household_wages, na.rm = TRUE))
```

This graphic shows the meidan (dot) and the 25th and 75th percentiles (linerange) of household income for each region of India.

```{r income_viz_box, fig.height=10}
household %>%
  mutate(stateid = fct_reorder(stateid, household_wages, median)) %>% 
  group_by(stateid) %>%
  summarise(p25 = quantile(household_wages, .25), median = median(household_wages), p75 = quantile(household_wages, .75)) %>%
  ungroup() %>%
  ggplot(aes(y = median/10000, x = stateid)) +
  geom_pointrange(aes(ymin = p25/10000, ymax = p75/10000)) +
# geom_vline(aes(xintercept = median), data = state_labels) +
  theme_bw() +
  labs(y = "Household income (10,000 INR)", x = "Region", title = "Distribution of Household Income") +
  theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))  

```

This [geography based facet](https://github.com/hafen/geofacet) of India provides histograms of the household income distribution by region in India.

```{r income_viz, fig.height=16, fig.width = 12}
household %>%
  ggplot(aes(x = household_wages/10000)) +
  geom_histogram() +
  theme_bw() +
  labs(x = "Household income (10,000 INR)", y = "Count", title = "Distribution of Household Income") +
  facet_geo(~stateid, grid = india_grid2, scales = "free") 

```


## Household Size

The household count numbers seem to make sense based on numbers reported at the following links. The median household size from this study data is `r round(median(household$n),2)`. 

- https://en.wikipedia.org/wiki/Indian_states_ranking_by_household_size
- http://www.arcgis.com/home/item.html?id=6cf22970ea8c4b338a196879397a76e4


```{r size_viz, fig.height=16, fig.width = 12}
household %>%
  ggplot(aes(x = n)) +
  geom_histogram(color = "white") +
# geom_vline(aes(xintercept = median), data = state_labels) +
  theme_bw() +
  labs(x = "Number in household", y = "Count", title = "Distribution of Household Size") +
  scale_x_continuous(breaks = seq(0,30, by = 5), limits = c(0, 15)) +
  theme(panel.grid.minor = element_blank()) +
  facet_geo(~stateid, grid = india_grid2,  scales = "free_y") 

```

```{r size_viz_box, fig.height=16, fig.width = 12}
household %>%
  mutate(stateid = fct_reorder(stateid, n, median)) %>% 
  group_by(stateid) %>%
  summarise(p25 = quantile(n, .25), median = median(n), p75 = quantile(n, .75)) %>%
  ungroup() %>%
  ggplot(aes(y = median, x = stateid)) +
  geom_pointrange(aes(ymin = p25, ymax = p75)) +
# geom_vline(aes(xintercept = median), data = state_labels) +
  theme_bw() +
  labs(y = "Number in household", x = "Region", title = "Distribution of Household Size") +
  scale_y_continuous(breaks = seq(0,30, by = 5), limits = c(0, 15)) +
  theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))  

```

## Under 5 years old height distributions

Some data quality issues will require further investigation. Also, the ages are grouped by year in the data.

```{r, fig.height=16, fig.width = 12}
# Household heights by age
kids %>%
  filter(age <= 5) %>%
  ggplot(aes(y = height2 * 0.0328084, x = factor(age)), fill = "darkgrey") +
  geom_boxplot() +
#  geom_jitter(aes(color = sex), alpha = .5, width = .25) +
  facet_geo(~stateid, grid = india_grid2) +
  theme_bw() +
  scale_color_brewer(type = "qual") +
  coord_cartesian(ylim = c(0, 4)) +
  labs(y = "height (ft)", x = "gender", color = "gender", title = "Distribution of height by age group")

```


## 0-1 year old heights by income category

Remember that this figure shows all newborns from 0-12 months.

```{r, fig.height=16, fig.width = 12}
kids %>%
  filter(age <= 1) %>%
  mutate(income_groups = 
           case_when( household_wages/ 10000 <= 2.5 ~ "less than 25k",
                      household_wages/ 10000 <= 5 ~ "between 25k & 50k",
                      household_wages/ 10000 <= 10 ~ "between 50k & 100k",
                      household_wages/ 10000 <= 25 ~ "between 100k & 250k",
                      household_wages/ 10000 > 25 ~ "greater than 250k")) %>%
  ggplot(aes(y =  height2 * 0.0328084, x = income_groups)) +
  scale_y_continuous(limits = c(0, 3)) + 
  geom_boxplot() +
  labs(x = "Household wages", y = "height (ft)") +
  coord_flip() +
  theme_bw()
```



```{r, fig.height=16, fig.width = 12}
kids %>%
  filter(age <= 1) %>%
  mutate(income_groups = 
           case_when( household_wages/ 10000 <= 2.5 ~ "less than 25k",
                      household_wages/ 10000 <= 5 ~ "between 25k & 50k",
                      household_wages/ 10000 <= 10 ~ "between 50k & 100k",
                      household_wages/ 10000 <= 25 ~ "between 100k & 250k",
                      household_wages/ 10000 > 25 ~ "greater than 250k")) %>%
  ggplot(aes(y =  height2 * 0.0328084, x = income_groups)) +
  scale_y_continuous(limits = c(0, 3)) + 
  geom_boxplot() +
  facet_geo(~stateid, grid = india_grid2) +
  labs(x = "Household wages", y = "height (ft)") +
  coord_flip() +
  theme_bw()
```

## Under 5 years old weight distributions

Some data quality issues will require further investigation.

```{r, fig.height=16, fig.width = 12}
kids %>%
  filter(age <= 5) %>%
  ggplot(aes(y = weight2 * 2.20462, x = factor(age)), fill = "darkgrey") +
  geom_boxplot() +
  #  geom_jitter(aes(color = sex), alpha = .5, width = .25) +
  facet_geo(~stateid, grid = india_grid2) +
  theme_bw() +
  scale_color_brewer(type = "qual") +
  coord_cartesian(ylim = c(0, 45)) +
  labs(y = "weight (lbs)", x = "gender", color = "gender", title = "Distribution of height by age group")


```

## 0-1-year-old weights by income category

Remember that this figure shows all newborns from 0-12 months.

```{r, fig.height=16, fig.width = 12}
kids %>%
  filter(age <= 1) %>%
  mutate(income_groups = 
           case_when( household_wages/ 10000 <= 2.5 ~ "less than 25k",
                      household_wages/ 10000 <= 5 ~ "between 25k & 50k",
                      household_wages/ 10000 <= 10 ~ "between 50k & 100k",
                      household_wages/ 10000 <= 25 ~ "between 100k & 250k",
                      household_wages/ 10000 > 25 ~ "greater than 250k")) %>%
  ggplot(aes(y = weight2 * 2.200462, x = income_groups)) +
  scale_y_continuous(limits = c(0, 25)) + 
  geom_boxplot() +
  labs(x = "Household wages (INR)", y = "weight (lbs)") +
  coord_flip() +
  theme_bw()
```



```{r, fig.height=16, fig.width = 12}
kids %>%
  filter(age <= 1) %>%
  mutate(income_groups = 
           case_when( household_wages/ 10000 <= 2.5 ~ "less than 25k",
                      household_wages/ 10000 <= 5 ~ "between 25k & 50k",
                      household_wages/ 10000 <= 10 ~ "between 50k & 100k",
                      household_wages/ 10000 <= 25 ~ "between 100k & 250k",
                      household_wages/ 10000 > 25 ~ "greater than 250k")) %>%
  ggplot(aes(y = weight2 * 2.200462, x = income_groups)) +
  scale_y_continuous(limits = c(0, 25)) + 
  geom_boxplot() +
  facet_geo(~stateid, grid = india_grid2) +
  labs(x = "Household wages (INR)", y = "weight (lbs)") +
  coord_flip() +
  theme_bw()
```


# 2011 IHDS Exploration (Births)

## Distribution of number of births per household

```{r, fig.height = 10}
dat <- read_rds(path = "../data/artifacts/2011_36151_India_births.Rds")


dat %>%
  group_by(idhh, stateid, mother_id) %>%
  summarise(births = max(nbirths)) %>%
  ungroup() %>%
  mutate(stateid = fct_reorder(stateid, births, quantile, probs = .75)) %>%
  ggplot(aes(x = stateid, y = births)) +
  geom_boxplot() + 
  coord_flip() +
  theme_bw() +
  labs(y = "Births per Mother", x = "Region")
```



# IHDS Data Prep

The development R package [tidyICPSR]() has the functions used to download and prep the data.

# Code

All of the R code to download, process, and replicate this analysis (including the code that generates this document) is available [here](https://github.com/hathawayj/ihds).



